\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2022-08-26 paper/si.tex
working on the manuscript

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

@@

\title{\bf  \sffamily \LARGE Evolution of cooperation among individuals with
limited payoff memory\\}
\date{}
\author{Nikoleta E. Glynatsi, Christian Hilbe, Alex McAvoy}

@@

\begin{abstract}
% Repeated games are vastly use in evolutionary game theory to explain cooperation in a variety of environments. Although existing models have shaped
% our understanding of human cooperation, hey often work with idealized
% assumptions. In an evolutionary process, individuals imitate other individuals
% of the population based on their fitness. It is commonly assumed that an
% individual computes their fitness after interacting with a representative sample
% of the population, and remembering all the interactions they participated in. In
% real life, we do not always remember all our interactions, instead we rather
% recall our most recent ones.

% Here, we introduce a framework that allows individuals to estimate their fitness
% based on a minimum of social information. We explore the difference between our
% framework and the classical framework using computer simulations. We present
% results for the most commonly used classes of symmetric \(2 \times 2\) games.
% The simulations show that, in the prisoner's dilemma individuals with limited
% memory tend to adopt less generous strategies and they achieve less cooperation
% than in the classical scenario. In contrast, in the stag-hunt and the harmony
% game, the impact of memory is less striking. Here individuals with limited
% memory perform nearly as well as individuals with full memory. Finally, we
% observe that the snowdrift game is the only game for which cooperation can be
% underestimated in the classical scenario.
\end{abstract}

\section{Introduction}

Evolutionary game theory~\cite{smith1982evolution, hofbauer1998evolutionary,
nowak2004evolutionary, hauert2005game} describes the evolutionary dynamics of
populations consisting of different types of interacting individuals. The
framework of evolutionary game theory has been applied in explaining the
behavior of sticklebacks~\cite{milinski1987tit}, the rock-scissors-paper cycles
in bacterial populations~\cite{kerr2002local} and food sharing in vampire
bats~\cite{carter2020development} colonies.

Traditional approaches of evolutionary game theory assume that individuals meet
each other at random in an infinitely large well-mixed population; an example of
such an approach is the replicator dynamics. The replicator dynamics describes
how the abundance of strategic types in a population changes based on their
fitness. In this deterministic formulation, individuals with higher fitness
increase in abundance and ultimately, the system reaches a stable fixed point in
which the population may consist either of a single type or of a mixture of
different types. The works of~\cite{nowak2004emergence, taylor2004evolutionary,
imhof2005evolutionary, imhof2006evolutionary, traulsen2007pairwise} have shown
that constraining the population to be finite can lead to fundamental changes
due to stochastic effects. Due to these stochastic effects disadvantageous
mutants have a small, yet non-zero probability to reach fixation in a finite
population.

Two classes of such finite stochastic processes have been used extensively: (i)
fitness-based processes in which an individual chosen proportional to fitness
reproduces and the offspring replaces a randomly chosen individual~\cite{nowak2004emergence}
\textit{moran process}; (ii) \textit{pairwise comparison processes} in which a
pair of individuals is chosen, and where subsequently one of these individuals
may adopt the strategy of the other~\cite{traulsen2007pairwise}.

A pairwise comparison process considers a finite population of fixed size. In
the population at a time step different types of individuals can exist. In the
simplest of cases each individual can be of one of two types, A and B. The state
of the population is thus characterized by the number \(k\) of individuals of
type A. The interaction between the two types of individuals is described by the
functions \(\pi _{{\rm A}}^{i}\) and $\pi _{{\rm B}}^{i}$.

At each step an individual change their type based. Two individuals, A and B,
are selected at random. The individual chosen for reproduction A replaces B with
probability \(\rho\), which depends on the fitness difference \(\pi _{{\rm
A}}^{i} - \pi _{{\rm B}}^{i}\) between the two individuals. The composition of
the population can only change if both individuals are of different types. The
probability \(\rho\) follows the Fermi function,

\begin{equation}
  \rho(\pi _{{\rm A}}^{i}, \pi _{{\rm B}}^{i}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (\pi _{{\rm A}}^{i}\!-\!\pi _{{\rm B}}^{i})}}.
\end{equation}

The parameter \(\beta\) denotes the intensity of selection.

The fitness of the individuals (\(\pi _{{\rm A}}^{i}, \pi _{{\rm B}}^{i}\)) is
identified with the payoff resulting from the underline game. The evolution of
cooperation remains one of the greatest problems for biological and social
sciences. Cooperation is the action of choosing to help others at one's own
expense. The standard game of formulating such situation is the Prisoner's
Dilemma~\cite{trivers1971evolution, milinski1987tit, glynatsi2021bibliometric,
ohtsuki2006simple}, in which two players can choose to cooperate or to defect.
The players are offered a certain payoff, \(R\), for mutual cooperation and a
lower payoff, \(P\), for mutual defection. If one player cooperates while the
other defects, then the cooperator gets the lowest payoff, \(S\), while the
defector gains the highest payoff, \(T\). Thus, the payoffs have the following
property \(T > R > P > S\), making defection the dominant strategy in the
non-repeated game.

\begin{equation}\label{eq:payoffs}
  \begin{blockarray}{ccc}
      & \text{cooperate} & \text{defect} \\
      \begin{block}{c(cc)}
          \text{cooperate} & R & S \\
          \text{defect} & T & P \\
      \end{block}
  \end{blockarray}
\end{equation}

In literature, the payoffs of A and B individuals only on the fraction of both
types in the population. If there are \(k\) A individuals and \(N-k\) B
individuals. In the context of the prisoner's dilemma the payoffs and assuming
that A and B follow simple strategies where they cooperate and defect
unconditionally respective, then the A and B individuals have fitness of

\begin{equation}
  \begin{array}{lcrcr}
  \displaystyle \pi _{{\rm A}}^{i}	&=	&\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot R	&+	&\displaystyle\frac{k}{N-1}\cdot S,\\[0.5cm]
  \displaystyle \pi _{{\rm B}}^{i}&=	&\displaystyle\frac{N-k}{N-1}\cdot T &+	&\displaystyle\frac{k-1}{N-1}\cdot P.\\
  \end{array}
\end{equation}

respectively while self-interactions are excluded. In literature these are the
expected payoffs. For \(\beta = 1\), fitness equals payoff. This scenario
describes ``strong selection''. For \(\beta \ll 1\) , the payoff only provides a
small perturbation to the overall fitness of an individual, a limit known as
weak selection.

In the evolutionary game theory literature it is usually assumed that players
use strategies with finite memory~\cite{Nowak1992tit, Baek2016}. This assumption
is common as it allows for an explicit calculation of the players'
payoffs~\cite{sigmund2010calculus}. Players' payoffs are calculated in the limit
of interactions assuming that the two strategies meet several times.

Customarily, most evolutionary game studies make the additional assumption that
individuals play many times and with all other players before reproduction takes
place, so that payoffs, equivalently fitness, are given by the mean distribution
of types in the population. This hypothesis, selection occurs much more slowly
than the interaction between individuals and more that remember all the
interactions they participated in. Although recent experimental studies show
that this may not always be the case in biology [12– 14], it is clear that in
cultural evolution or social learning the time scale of selection is much closer
to the time scale of interaction. The effects of this mixing of scales cannot be
disregarded [15], and then it is natural to ask about the consequences of the
above assumption and the effect of relaxing it. The work of~\cite{Roca2006}.

In this work we ask studies that make use of this classical framework report
that cooperation can substantially evolve. However, the inconsistency in the
memory size of players leads us to question the robustness of our understanding
of cooperation. To this end, we propose a framework in which individuals,
similar to the decisions at each turn, estimate their fitness based on a minimum
of information.

We first consider two extreme scenarios, the classical scenario and the
alternative scenario where individuals update their strategies only based on the
very last payoff they obtained. We observe that individuals with limited memory
tend to adopt less generous strategies and they achieve less cooperation when
interacting in a prisoner's dilemma. We obtain similar results when we consider
that individuals update their strategies based on more information. More
specifically, up to the last two payoffs they obtained when interacting with up
to two different members of the population. We extend our approach to the rest
of the symmetric \(2 \times 2\) games.

The remainder of the paper is organized as follows. In
section~\ref{section:model} we describe the model. In
section~\ref{section:results} we present the results of the simulations, and in
section~\ref{section:conclusions} we outline the main conclusions.

\section{Model Setup}\label{section:model}

In the following, we consider a well mixed population of fixed size. In each
step of t

We study the transmission of strategies with a frequency-dependent birth-death
process26 in a finite population of size n. In each time step, two randomly
chosen individuals compare their payoffs and one of them can switch to the other
one's strategy. This process can be interpreted as a model for social learning,
whereby successful strategies spread, and, occasionally, random strategy
exploration introduces novel strategies (corresponding to mutations in
biological models).

We consider a population of \(N\) players~\footnote{The terms ``player'' and
``individual'' are used interchangeably here.} where \(N\) is even and mutations
are sufficiently rare. Therefore, at any point in time there are at most two
different strategies present in the population; a \textit{resident} strategy and
a \textit{mutant} strategy. To describe how strategies spread we use a pairwise
comparison process~\cite{Traulsen2006}. Each step of the evolutionary process
consists of two stages, a game stage and an updating stage.

In the game stage each individual is randomly matched with some other individual
in the population. They engage in a match where each subsequent turn occurs with
a fixed probability $\delta$. At each turn players choose independently to
either cooperate (\(C\)) or to defect (\(D\)), and the payoffs of the turn
depend on both their decisions. If both players cooperate they receive the
reward payoff \(R\), whereas if both defect they receive the punishment payoff
\(P\). If one cooperates but the other defects, the defector receives the
temptation payoff \(T\), whereas the cooperator receives the sucker's payoff
\(S\). We denote the feasible payoff of each turn as \(\mathcal{U} = \{R, S, T,
P\}\). We assume that individuals use \textit{reactive strategies} to make
decisions in each turn. Reactive strategies are a set of memory-one strategies
that only take into account the previous action of the opponent. They can be
written explicitly as a vector in \(\R_{3}\), more specifically, a reactive
strategy \(s\) is given by \(s=(y, p, q)\). The parameter \(y\) is the
probability that the strategy opens with a cooperation and \(p, q\) are the
probabilities that the strategy cooperates given that the opponent cooperated
and defected equivalently.

In the updating stage, two players are randomly drawn from the population, a
`learner' and a `role model'. Given the learner's payoff $u_L\!\in\!
\mathcal{U}$ and the role model's payoff $u_{RL}\!\in\! \mathcal{U}$, the
learner adopts the role model's strategy with probability,

\begin{equation} \label{Eq:rho}
  \rho(u_{L}, u_{RM}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (u_{RM}\!-\!u_{L})}}.
\end{equation}

where $\beta\!\ge\!0$ is the strength of selection. For small values of $\beta$
the imitation probability is independent of the strategies of the involved
players. As the value of $\beta$ increases, the more likely it is that the
learner adopts only strategies that yield a higher payoff. Conventionally the
updating payoffs of the learner and the role model are based on their expected
payoffs. A player's expected payoff is the mean payoff the player yields after
engaging in matches of multiple turns with each member of the population. At
each match a player bases their next turn decision only on the previous action
of the opponent, however, the same player bases their expected payoffs on the
outcomes of all their matches. Thus, a player is assumed to have limited and
perfect memory at the same time. We propose a new a set of updating payoffs
where it is also assumed that a player has limited memory. We referee to these
as the limited memory payoffs.

The evolutionary step is repeated until either the mutant strategy goes extinct,
or until it fixes in the population. If the mutant fixes in the population then
the mutant strategy becomes the new resident strategy. After either outcome we
introduce a new mutant strategy uniformly chosen from all reactive strategies at
random, and we set the number of mutants to $1$. This process of mutation and
fixation/extinction is then iterated many times.

In order to account for the effect of the updating payoffs we simulate the
evolutionary process and record which strategies the players adopt over time
based on (i) the expected payoffs (ii) the limited memory payoffs. We compare
the difference in the cooperation rate within the resident population for the
two approaches. To account for the various types of social behaviour we also
present results on multiple social dilemmas.

\section{Results}\label{section:results} \subsection{Updating payoffs based on
the last round with another member of the population}\label{section:donation}

To investigate the role of extortion in the context of evolutionary games, we
concentrate on the donation game.
In this section we explore the case where the updating payoffs are based on the
last round payoff achieved against another member of the population and we
compare this to the classical scenario of the expected payoffs. We assume that
each pair of players interacts in a donation game. The donation game is a
special case of the prisoner's dilemma. Each player can choose to cooperate by
providing a benefit \(b\) to the other player at their cost \(c\), with \(0 < c
< b\). Thus, the feasible payoffs in each round are \(\mathcal{U} = \{b-c, -c,
b, 0\}\).

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulations results
for the described process of section~\ref{section:model}.
Figure~\ref{fig:expected_and_stochastic_for_donation} depicts the evolving
conditional cooperation probabilities $p$ and $q$. The discount factor~$\delta$
is comparably high, thus we do not report the opening move \(y\) as it is a
transient effect. The left panels correspond to the standard scenario considered
in the literature, it considers players who use expected payoffs to update their
strategies. The right panel shows the scenario considered herein, in which
players update their strategies based on their last round’s payoff. The top
panels assume a benefit \(b\) of 3 whereas the bottom assume a benefit of 10.

The figure suggests that when updating is based on expected payoffs players tend
to be more generous and more cooperative. The $q$-values of the resident
strategies are on average higher in the case of the expected payoffs. The
players will occasionally forgive a defection more often if their fitness
depends on interacting with every member of the population. On the other hand,
when social interactions are limited they are less forgiving. The average
cooperation rate for each simulation is calculated as the average cooperation
rate within the resident population. In the case of the expected payoffs,
regardless the value of benefit, the average cooperation rate is strictly higher
than that of the last round payoffs. The difference based on the two methods is
statistically significant, and in the case of $b=10$ the average cooperation of
resident strategies drops from 97\% to 57\%.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.70\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}
    % \includegraphics[width=.70\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and last round with one interaction payoffs.} 
    We have run two simulations of the evolutionary process described in
    section~\ref{section:model} for $t\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since
    simulations are run for a relatively high continuation probability of
    $\delta\!=\!0.999$, we do not report the players' initial cooperation
    probability $y$. The graphs show how often the resident population chooses
    each combination ($p,q$) of conditional cooperation probabilities in the
    subsequent rounds. ({\bf A}) If players update based on their expected
    payoffs, the resident population typically applies a strategy for which
    $p\!\approx\!1$ and $q\!\le\!1\!-\!c/b\!=\!0.9$. ({\bf B}) When players
    update their strategies based on their realized payoffs in the last round,
    there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$
    consistent with stable cooperation is somewhat smaller compared to the
    expected-payoff setting, $q\!<\!0.5$. The cooperation rate within the
    resident population (averaged over all games and over all time steps) is
    close to 100\%. Parameters: $N\!=\!100$, $c\!=\!1$, $\beta\!=\!1$,
    $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

\begin{figure}[!htbp]
  \centering
  \includegraphics[width=.9\textwidth]{static/cooperation_rates_expected_and_stochastic_for_donation_game.pdf}
\end{figure}

We further explore the effect of benefit in
Figure~\ref{fig:cooperation_rate_over_benefit}. The figure suggests that
expected payoffs always yield a higher cooperation rate. In the case of expected
payoffs we observe that the cooperation rate increases as the value of the
benefit gets higher. In comparison for the limited memory payoffs, the
cooperation rate remains unchanged at approximately 50\% once \(b=5\).

% \begin{figure}[!htbp]
%   \centering
%   \begin{subfigure}{.5\textwidth}
%     \centering
%     \includegraphics[width=\textwidth]{static/expected_for_beta.pdf}
%     \includegraphics[width=\textwidth]{static/stochastic_for_beta.pdf}
%   \end{subfigure}%
%   \begin{subfigure}{.5\textwidth}
%     \centering
%     \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
%   \end{subfigure}
%   \caption{{\bf The evolution of cooperation for different benefit values.} 
%   We vary the benefit of defection $b$. In all cases, expected payoffs appear to
%   overestimate the average cooperation rate the population achieves. ({\bf A})
%   the probabilities \(p, q\) for resident population over \(10^7\) time steps
%   for each benefit value. ({\bf B}) The cooperation rate within the resident
%   population (averaged over all games and over all time steps) over the benefit.
%   Unless explicitly varied, the parameters of the simulation are $N\!=\!100$,
%   $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
%   $t\!=\!5\times 10^7$ time steps for each parameter
%   combination.}\label{fig:cooperation_rate_over_benefit}
% \end{figure}

We also investigate the effect of the strength of selection $\beta$.
Figure~\ref{fig:cooperation_rate_over_betas} illustrates results for various
runs of the evolutionary process. For weak selection, \(\beta < 1\), we observe
that the two methods yield similar results, however, as \(\beta\) increases
there is variation in the evolving populations. In the case of expected payoffs
the resident populations become more cooperative as \(\beta\) increases, whereas
in the case of limited memory payoffs, the resident populations become more
defective.

\begin{figure}[!htbp]
  \centering
    \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
\caption{{\bf The evolution of cooperation for different selection strength values.}
We vary the selection strength $\beta$. In all cases, stochastic payoff
evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
probabilities \(p, q\) for resident population over \(10^7\) time steps for each
\(\beta\) value. ({\bf B}) The cooperation rate within the resident population
(averaged over all games and over all time steps) over \(\beta\). Unless
explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$,
$c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
$t\!=\!5\times 10^7$ time steps for each parameter
combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

\begin{figure}[!htbp]
  \centering
    \includegraphics[width=\textwidth]{static/more_memory.pdf}
% \caption{{\bf The evolution of cooperation for different selection strength values.}
% We vary the selection strength $\beta$. In all cases, stochastic payoff
% evaluation tends to reduce the evolving cooperation rates. ({\bf A}) the
% probabilities \(p, q\) for resident population over \(10^7\) time steps for each
% \(\beta\) value. ({\bf B}) The cooperation rate within the resident population
% (averaged over all games and over all time steps) over \(\beta\). Unless
% explicitly varied, the parameters of the simulation are $N\!=\!100$, $b\!=\!3$,
% $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.99$. Simulations are run for
% $t\!=\!5\times 10^7$ time steps for each parameter
% combination.}\label{fig:cooperation_rate_over_betas}
\end{figure}

% \subsection{Effect of updating payoffs in different social
% dilemmas}\label{section:2_by_2_games}

% In the previous section we gained insights into the effects of the updating
% payoffs, and into how parameters such as the benefit and the strength of
% selection can intensify them. We investigated these effects by using the
% donation game. In order to broaden our understanding of the updating payoffs on
% different forms of possible human interactions we extend our approach to all \(2
% \times 2\) symmetric games. More specifically, we apply our analysis to four
% different classes of games; the harmony, the snowdrift, the prisoner's dilemma
% and the stag-hunt games.

% We compare the results of the evolutionary process when the updating payoffs are
% based on the expected payoffs, and on the last round payoffs for all the four
% possible classes of games (Figures~\ref{fig:expected_two_by_two}
% and~\ref{fig:last_round_two_by_two}). In Figure~\ref{fig:expected_two_by_two}
% each sub figure represents a run of evolutionary process for a different set of
% values for \(\mathcal{U}\). Without loss of generality we set \(R=1\) and
% \(P=0\)~\cite{Martinez2012, Roca2009}, and we vary the values of the temptation
% payoff \(T\) (across the \(x\)-axis) and of the sucker's payoff \(S\) (across
% the \(y\)-axis). Starting at the upper left corner and proceeding clockwise the
% quadrants correspond to the harmony, the snowdrift, the prisoner's dilemma and
% the stag-hunt games.

% \begin{figure}[!htbp]
%   \centering
%   \includegraphics[width=\textwidth]{static/expected_two_by_two_games.pdf}
%   \caption{{\bf Evolutionary dynamics under expected payoffs for various social dilemmas.}
%   We vary the temptation payoff \(T \in \{0.2, 0.6, 1, 1.4, 1.8, 2.2\}\) across
%   the \(x\) axis, and  \(S \in \{1.2, 0.8, 0.4, 0, -0.4, -0.8\}\) across the
%   \(y\) axis. ({\bf A}) The top left quadrant where \(S > 0\) and \(T \leq 1\)
%   corresponds to the harmony game. The preference ordering for the harmony game
%   is \(R > T > S > P\). ({\bf B}) The top right quadrant where \(S > 0\) and \(T
%   > 1\) corresponds to the snowdrift game. The preference ordering for the
%   snowdrift game is \(T > R > S > P\). ({\bf C}) The bottom left quadrant where
%   \(S < 0\) and \(T \leq 1\) corresponds to the stag-hunt game. The preference
%   ordering for the stag-hunt game is \(R > T > P > S\). ({\bf D}) The bottom
%   right quadrant where \(S < 0\) and \(T > 1\) corresponds to the prisoner's
%   dilemma game. The preference ordering for the prisoner's dilemma game is \(T >
%   R > P > S\). Unless explicitly varied, the parameters of the simulation are
%   $N\!=\!100$, $\beta\!=\!10$, $\delta\!=\!0.99$, $t\!=\!5\times 10^7$.}
%   \label{fig:expected_two_by_two}
% \end{figure}

% \begin{figure}[!htbp]
%   \centering
%   \includegraphics[width=\textwidth]{static/stochastic_two_by_two_games.pdf}
%   \caption{{\bf Evolutionary dynamics under last round payoffs for various social dilemmas.} 
%   We vary the temptation payoff \(T \in \{0.2, 0.6, 1, 1.4, 1.8, 2.2\}\) across
%   the \(x\) axis, and  \(S \in \{1.2, 0.8, 0.4, 0, -0.4, -0.8\}\) across the
%   \(y\) axis. ({\bf A}) The top left quadrant where \(S > 0\) and \(T \leq 1\)
%   corresponds to the harmony game. The preference ordering for the harmony game
%   is \(R > T > S > P\). ({\bf B}) The top right quadrant where \(S > 0\) and \(T
%   > 1\) corresponds to the snowdrift game. The preference ordering for the
%   snowdrift game is \(T > R > S > P\). ({\bf C}) The bottom left quadrant where
%   \(S < 0\) and \(T \leq 1\) corresponds to the stag-hunt game. The preference
%   ordering for the stag-hunt game is \(R > T > P > S\). ({\bf D}) The bottom
%   right quadrant where \(S < 0\) and \(T > 1\) corresponds to the prisoner's
%   dilemma game. The preference ordering for the prisoner's dilemma game is \(T >
%   R > P > S\). Unless explicitly varied, the parameters of the simulation are
%   $N\!=\!100$, $\beta\!=\!10$, $\delta\!=\!0.99$, $t\!=\!5\times 10^7$.}
%   \label{fig:last_round_two_by_two}
% \end{figure}

% The harmony game describes an interaction without conflict. In the harmony game
% it is in the best interest for both players to cooperate, and in an evolutionary
% process cooperators prevail. We observe that for most harmony game runs the
% resident population overwhelmingly applies a strategy for which \(p\) and \(q\)
% are \(\approx 1\) (Figure~\ref{fig:expected_two_by_two}). The snowdrift game
% describes a situation similar to that of the prisoner's dilemma; cooperation
% results in a benefit to the opposing player but entailing a cost to the
% cooperator. However, in the snowdrift game individuals obtain immediate direct
% benefits from the cooperative acts which leads to \(S > P\). In the snowdrift
% game more cooperation emerges compared to the prisoner's dilemma. Defection is
% never a resident strategy and for the same values of temptation the overall
% \(q-\) values are higher. The last class of games we present are for the
% stag-hunt game. In the stag-hunt game there is an equilibrium in which both
% players cooperate as well as one in which both defect. For small values of the
% temptation and the sucker's payoffs we observe that cooperation prevails similar
% to the harmony game. However, in the case where the temptation and the reward
% payoffs are equal the resident population either consists of defectors or
% cooperators.

% Although the patterns of the evolved populations remain similar if we consider
% that players update their strategies based on their last round payoff
% (Figure~\ref{fig:last_round_two_by_two}), there are still dissimilarities
% between the two approaches. To better understand these dissimilarities, we
% explore the cooperation rates for each of the cases of the social dilemmas we
% have presented (Figure~\ref{fig:cooperation_rates_two_by_two_part_one}).

% In the harmony game the cooperation rates for both approaches are similar. The
% biggest difference occurs for \(T=1\), nevertheless, it is less than 10\%. There
% are two instances for which the cooperation rates are higher for the limited
% memory payoffs compared to the expected. These are for \(T=2.2, S=0.4\) and for
% \(T=2.2, S=0.8\). It is only in the snowdrift game that cooperation is
% overestimated when a minimum of information is considered. Players are slightly
% more likely to cooperate after a cooperation and to forgive after being at the
% receiving end of a defection. For the rest of the cases the classical scenario
% overestimates the average cooperation rate, supporting the results of the
% previous section. There is a substantial difference in the cooperation rates for
% the prisoner's dilemma class. We observe that for \(S<-0.4\) cooperation almost
% never emerges if players imitate based only on the last round payoff.

% \begin{figure}[!htbp]
%   \centering
%   \includegraphics[width=.4\textwidth]{static/expected_two_by_two_games_cooperation.pdf}
%   \includegraphics[width=.4\textwidth]{static/stochastic_two_by_two_games_cooperation.pdf}
%   \caption{\textbf{Cooperation rates for various social dilemmas.} ({\bf A}) If
%   players update based on their expected payoffs. ({\bf B}) If players update
%   based on their last round payoffs. We vary the temptation payoff \(T \in
%   \{0.2, 0.6, 1, 1.4, 1.8, 2.2\}\) across the \(x\) axis, and \(S \in \{1.2,
%   0.8, 0.4, 0, -0.4, -0.8\}\) across the \(y\) axis. Unless explicitly varied,
%   the parameters of the simulation are $N\!=\!100$, $\beta\!=\!10$,
%   $\delta\!=\!0.99$, $t\!=\!5\times
%   10^7$.}\label{fig:cooperation_rates_two_by_two_part_one}
% \end{figure}

% So far we have explored the difference between the expected payoffs and the last
% round payoffs. In order to explore further the effect of limited memory we allow
% individuals to remember more. More precisely, up to two interactions and up to
% the last two rounds. In total we present results for three more updating
% payoffs. These are the payoffs when individuals consider the last two rounds
% with another member of the population, the last round with two members of the
% population, and the last two rounds with two members of the population. Similar
% to the last round payoff, we use simulations and record which strategies the
% players adopt over time based
% (Appendix~\ref{appendix:further_simulation_results}) and we compare the evolving
% cooperation rates to those of the expected payoffs
% (Figure~\ref{fig:cooperation_rates_two_by_two_part_two}).

>>>>

\documentclass[11pt]{article}
\usepackage[dvipsnames]{xcolor}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{blkarray}
\usepackage[ruled,vlined]{algorithm2e}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

\usetikzlibrary{decorations.pathreplacing, backgrounds, fit, calc, matrix, positioning}

@@

\definecolor{expectedcolor}{rgb}{0.1791464821222607, 0.49287197231833907, 0.7354248366013072}
\definecolor{oneone}{rgb}{0.8503344867358708, 0.14686658977316416, 0.13633217993079583}
\definecolor{onetwo}{rgb}{0.18246828143021915, 0.5933256439830834, 0.3067589388696655}
\definecolor{twoone}{rgb}{0.4488427527873895, 0.3839600153787005, 0.6738792772010764}
\definecolor{twotwo}{rgb}{0.8871510957324106, 0.3320876585928489, 0.03104959630911188}

@@

\title{\bf  \sffamily \LARGE Supplementary Information: Evolution of cooperation among individuals with
limited payoff memory\\}
\date{}
\author{Nikoleta E. Glynatsi, Christian Hilbe, Alex McAvoy}

@@

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. Section~\ref{section:updating_payoffs} describes the
conventional approach for calculating updating payoffs, as well as, our
approaches and methods.

\section{Pairwise comparison process}\label{section:pairwise_comparison}

Pairwise comparison process is a stochastic process for modelling the evolution
of a finite population. The process starts with assigning all individuals of the
population the same strategy. A strategy is a set of rules of how an individual
should behave in an interaction with another individual. Each elementary time
step of the process consists of three phases; (1) the \textbf{mutation phase}
(2) the \textbf{game phase} (3) the \textbf{update phase}. These are summarised
in Figure~\ref{fig:pairwise_phases}. In the mutation phase one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\). The mutant
strategy is randomly selected from the set of feasible strategies.

In the game phase individuals are randomly matched with other individuals in the
population, and they engage in a match where each subsequent turn occurs with a
fixed probability \(\delta\). At each turn the individuals play a social dilemma
and decide on an action based on their strategies. In the donation game there
are two actions: cooperation (\(C\)) and defection (\(D\)). By cooperating a player
provides a benefit \(b\) to the other player at their cost \(c\), with \(0 < c <
b\). Thus the payoffs for a player in each turn are,

\begin{equation}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & b - c & -c \\
            \text{defect} & b & 0 \\
        \end{block}
    \end{blockarray}.
\end{equation}

Let \(u = (b-c, -c, b, 0)\) be payoffs in a vector format, and let \(\mathcal{U}
= \{b-c, -c, b, 0\}\) denote the set of feasible payoffs. In repeated games
there are infinite many strategies. However, it is commonly assumed that
individuals can only choose strategies from a restricted set. Here we explore
the case where individuals can use reactive strategies. A reactive strategy
considers only the previous action of the other player, and thus, a reactive
strategy \(s\) can be written as a three-dimensional vector \(s=(y, p, q)\). The
parameter \(y\) is the probability that the strategy opens with a cooperation
and \(p\), \(q\) are the probabilities that the strategy cooperates given that
the opponent cooperated and defected equivalently.

Following the game phase is the update stage where two individuals are randomly selected.
One individual serves as the `learner' and the other as the `role model'.
The learner adopts the role model's strategy with a probability \(\rho\) given by,

\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{RM}) = \frac{1}{1\!+\! \exp^{\!-\!\beta (\pi_{RM}\!-\! \pi_{L})}}.
\end{equation}

\(\pi_{L}\) and \(\pi_{RM}\) are the updating payoffs of the learner and the
role model respectively. The updating payoff is a measure of how successful an
individual is in the current standing of the population. In the next section we
will explain how these payoffs are calculated in more detail. The parameter
\(\beta\) is known as the selection strength, namely, it shows how important the
payoff difference is when the learner is considering adopting the strategy of
the role model.

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise process phases.} \textbf{A)} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new mutant strategy. \textbf{C) Game phase.} individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(delta\) that a
    subsequent turn will occur, and with a probability \(1- delta\) the interaction ends.
    \textbf{D) Updating phase.} At the updating phase two individuals are chosen;
    one is assigned the role of the learner and the other one the role
    of the role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(pi_{C}, \pi_{D}\) are the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

This elementary population updating process is repeated for a large number of
time steps, and at each time step we record the state of the population.

\subsection{Low mutation \(\mu \rightarrow 0\)}

In the case of low mutation (\(\mu \rightarrow 0\)) we assume that mutations are
rare. In fact, so rare that only two different strategies can be present in the
population at any given time. The case of low mutation is vastly adopted because
it allows us to explicitly calculate the fixation probability of a newly
introduced mutant.

More specifically, the process again starts with a population where all members are of the same
strategy. At each step one individual adopts a mutant strategy randomly selected
from the set of feasible strategies. The fixation probability \(\phi_{M}\) of the
mutant strategy can be calculated explicitly,
